---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# psc

<!-- badges: start -->
  [![R-CMD-check](https://github.com/richJJackson/pic/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/richJJackson/pic/actions/workflows/R-CMD-check.yaml)
[![R-CMD-check](https://github.com/richJJackson/pic/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/richJJackson/pic/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of psc is to compare a dataset of observations against a parametric model

## Installation

You can install the development version of psc from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("richJJackson/psc")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(psc)
library(survival)
## basic example code

### Load model
data("surv.mod")

### Load Data
data("data")

### Use 'pscfit' to compare
surv.psc <- pscfit(surv.mod,data)

```

You can use standard commands for getting a summary of your analysis...
```{r cars}
summary(surv.psc)
```

... and to see a plot of what you have done

```{r pressure, echo = FALSE}
plot(surv.psc)
```

In that case, don't forget to commit and push the resulting figure files, so they display on GitHub and CRAN.
